---
title: "Generate Figure 1 and Figure 2A"
output: html_document
date: "2023-06-22"
---
```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```
## Setup
Let's start by loading in our libraries, setting our working directory, and sourcing some functions
```{r}
library(ggplot2)
library(ggsignif)
library(graphics)
library(matlab)
library(DBI)
library(xtable)
library(egg)
library(AUCell)
library(xlsx)
library(growthrates)
library(purrr)
library(ggbreak)
library(dplyr)
library(tidyr)
setwd('~/Repositories/densityDependentSelection/code') 
source('getGrowthModel.R')
source('utils.R')
```

## Get Cell Line r/K Parameters
This section does not produce any figures, but it produces the list 'fits' which will be used further down. The required input to this section is a list of cell line lineages with matching growth data .txt files in data/cellCounts. The functional script getGrowthModel.R uses the R package 'growthrates' to perform nonlinear regression and return the inferred r/K parameters and summary statistics which are gathered into the 'fitSummaryTable' data frame.
```{r}
lineages = sort(c("NUGC-4_A4_seed","NCI-N87_A59_seed","HGC-27_A22_seed","KATOIII_A6_seed","SNU-668_A9_seed" ,"MKN-45_A4_seed","SNU-638_A4_seed","SNU-601_A4_seed"));
cellLines = gsub("\\_.*","",lineages)
cellLinePairs = t(combn(unique(cellLines), 2))
modelTypes = list(Logistic = grow_logistic)
rSquares = matrix(NA, length(modelTypes), length(lineages), dimnames = list(names(modelTypes), lineages))
fits = list()
for (lineage in lineages) {
  tmp =try(getGrowthModel(lineage,modelTypes[["Logistic"]]))
  tmp$type = "Logistic"
  fits[[lineage]] = tmp
  targetDate = sort(as.POSIXct(tmp$out$date), decreasing=F)[round(nrow(tmp$out)*0.25)]
  rSquares["Logistic", lineage] = tmp$adjr2
}
fitSummaryTable = data.frame(matrix(NA, nrow = length(lineages), ncol = 3, dimnames = list(lineages, c("r", "K", "adjutedRSquared"))))
fitSummaryTable$adjutedRSquared=sapply(lineages, function(x) round(max(rSquares[,x]), 4))
fitSummaryTable$r=sapply(lineages, function(x) round(fits[[x]]$modelsum@par["mumax"],2))
fitSummaryTable$K=sapply(lineages, function(x) round(fits[[x]]$modelsum@par["K"]))
fitSummaryTable$cellLine = gsub("\\_.*", "", rownames(fitSummaryTable))
rownames(fitSummaryTable) = fitSummaryTable$cellLine
fitSummaryTable[,1:3]
```

## Generate Figure 1A
Next, we gather up all the individual fits per cell line into one data frame and plot together. We also shift their curves along the x-axis such that the inflection point occurs at x=0 and plot in log-scale.
```{r}
# Find x-axis range need to accommodate all fits
maxDay = max(unlist(sapply(fits, function(x) x$modelsum@obs$time)))
x = seq(0, maxDay, by = 0.01)
# Extract our y-values from fits
ggplotGrowthCurves = lapply(fits, function(v) y = (v$modelsum@par["K"] * v$modelsum@par["y0"])/(v$modelsum@par["y0"] + (v$modelsum@par["K"] - v$modelsum@par["y0"]) * exp(-v$modelsum@par["mumax"] * x)))

# Shift plots along the x-axis so that they cross x=0 at their max growth rate
medians = sapply(ggplotGrowthCurves, function(x) median(0:max(na.omit(x))))
diffVals = sapply(names(fits), function(x) abs(ggplotGrowthCurves[[x]] - medians[[x]]))
daysToSubtract = apply(diffVals, 2, function(v) x[which.min(v)])

# Setup colors so that it is consistent across all plots
dd.col = RColorBrewer::brewer.pal(length(cellLines),"Paired");
names(dd.col) = cellLines

# Grab (x,y) coords for all cell line growth data and store in 'ggplotGrowthDataPoints'
ggplotGrowthDataPoints = list()
for (name in names(fits)){
  ggplotGrowthDataPoints[[name]] = fits[[name]]$modelsum@obs$time-daysToSubtract[[name]]
}
ggplotGrowthDataPoints = round(as.data.frame(unlist(ggplotGrowthDataPoints)),2)
names(ggplotGrowthDataPoints) = "xPoints"
ggplotGrowthDataPoints$yPoints = unlist(lapply(names(fits), function(x) fits[[x]]$modelsum@obs$y))
ggplotGrowthDataPoints$CL = gsub("\\_.*","",rownames(ggplotGrowthDataPoints))

# Rename daysToSubtract to match cell lines and populate ggplotGrowthCurves with shifted x-values (ggplotGrowthCurves$newDays)
names(daysToSubtract) = gsub("\\_.*","",names(daysToSubtract))
ggplotGrowthCurves = as.data.frame(unlist(ggplotGrowthCurves))
names(ggplotGrowthCurves) = "y"
ggplotGrowthCurves$CL = gsub("\\_.*","",rownames(ggplotGrowthCurves))
ggplotGrowthCurves$x = repmat(x,1)
ggplotGrowthCurves$newDays=unlist(lapply(unique(ggplotGrowthCurves$CL), function(z) ggplotGrowthCurves[ggplotGrowthCurves$CL==z, "x"]-daysToSubtract[z]))

# here we adjust the ggplotGrowthCurves so that each curve will end at the last data point for that cell line
# above when we find the ggplotGrowthCurves they are all solved over same timespan - this loop gives them all unique x-axes
adjuster = list()
for(CL in cellLines){
  cutOff = max(ggplotGrowthDataPoints[ggplotGrowthDataPoints$CL==CL, "xPoints"])
  tmp = ggplotGrowthCurves[ggplotGrowthCurves$newDays<=cutOff, ]
  tmp = tmp[tmp$CL==CL,]
  adjuster[[CL]] = tmp
}

ggplotGrowthCurves = adjuster %>% map_df(~rbind(.x), .id = 'CL')

# Create data frame to pass through to ggplot that contains ggplotGrowthCurves and ggplotGrowthDataPoints
dfT = list(df1 = ggplotGrowthCurves, df2 = ggplotGrowthDataPoints) %>% 
  map_df(~rbind(.x), .id = 'df')

ggplot(data=dfT, aes(x=newDays, y=y, color=CL)) + 
  geom_line(linewidth=2) + geom_point(aes(x=xPoints, y=yPoints), size=4) + scale_color_manual(values=dd.col) + 
  scale_y_log10() + xlab("Time (days)") + ylab("Cell count") + theme_minimal() + 
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 18), legend.position="none")
```

## Generate Figures 1C-D,1F-G
Here we go through KEGG pathways identified as differentially expressed between r-selected and K-selected HeLa cells and see which are the best biomarkers for these parameters in our cell lines. This section requires combined Seurat objects (i.e., all cell lines in one Seurat object) per cell cycle state in data/SeuratObjects and will create AUCell objects to be stored in data/auCellObjects if they are not alreay present in that directory. We also print the predicted r/K params for each subpopulation per cell line in this section.
```{r}
# If no AUCell objects in directory, run AUCell on Seurat objects
if(length(list.files("../data/auCellObjects"))<4){
  for(state in c("S","G0G1","G2M","all")){
    uList = getTranscriptomeData(cellLines, ccstate = state)
    save(file=paste0("../data/auCellObjects/uList_",state,".RObj"), "uList")
  }
}

# Set up list containing a pathway per r/K parameter
inputList = list("K"= "Arginine and proline metabolism", "r" = "Amoebiasis")
axesTitleList = list("K"= "Carrying capacity (K)", "r" = "Growth rate (r)")

# Compare pathways from hela paper to growth parameters:
poi=read.xlsx("../data/pathways/TableS2andS3_HelaPaper.xlsx",sheetName =  "LowDensity")
subpopGrowthParams = list()
pairs = teList = cellCycleClones = list()
maxPval=0.1
mapSigLevel = c("***"=0.001, "**"=0.01, "*"=0.05, "o"=0.1)
for(param in names(inputList)){
  teList_ = list()
  for(state in c("S","G0G1","G2M", "all")){
    # load CC specific AUCell object:
    load(file=paste0("../data/auCellObjects/uList_",state,".RObj"))
    teList_[[state]]=getStats4ParamFit(param,uList,poi, verbose=F)$te
    if(state=="G0G1"){
      cloneMembershipBarplot = as.data.frame(plyr::count(paste(uList$cloneMembership, uList$clMembership)))
      cloneMembershipBarplot = cloneMembershipBarplot %>% separate(x, c('clone', 'cellLine'), sep = " ")
      rownames(cloneMembershipBarplot) = cloneMembershipBarplot$clone
      cloneMembershipBarplot$ratio=sapply(cloneMembershipBarplot$clone, function(x) cloneMembershipBarplot[cloneMembershipBarplot$clone==x, "freq"]/sum(cloneMembershipBarplot[cloneMembershipBarplot$cellLine==cloneMembershipBarplot[cloneMembershipBarplot$clone==x,"cellLine"],"freq"]))
    }
  }
  teList[[param]]=do.call(rbind,teList_)
  # Choose cell cycle state with best model performance for this param:
  teList_=sapply(teList_, function(x) x[1,])
  state=colnames(teList_)[which.max(teList_[1,])]
  load(file=paste0("../data/auCellObjects/uList_",state,".RObj"))
  out =getStats4ParamFit(param,uList,poi)
  model = out$model
  
  # Build mapping of clones to cell lines
  clonesToCellLinesMap = as.data.frame(plyr::count(paste(uList$cloneMembership, uList$clMembership)))
  clonesToCellLinesMap = clonesToCellLinesMap %>% separate(x, c('clone', 'cellLine'), sep = " ")
  rownames(clonesToCellLinesMap) = clonesToCellLinesMap$clone
  
  # model predictions
  scDat=data.frame(indepVar1 = uList$pathways[inputList[[param]],])
  predVar = predict(model, scDat)
  if(param == "K"){
    predVar = exp(predVar)
  }
  
  ## set up data frame to use in ggplot
  scDat$prediction = predVar
  scDat$clone = uList$cloneMembership
  subpopGrowthParams[[param]] = as.data.frame(grpstats(as.matrix(scDat$prediction), as.matrix(scDat$clone), "median"))
  scDat$cellLine = uList$clMembership
  
  # get plot order for clones
  scDat = getPathwayActivityRanksAcrossCellLines(scDat)
  compsList = getLargestSignificantPair(scDat, clonesToCellLinesMap,cellLinePairs,maxOnly = T, maxPval = maxPval)
  list2env(compsList, .GlobalEnv)
  remove(compsList)
  
  # set step increase for plotting
  sI = 0.08
  mT = 0.005
  if(param=="K"){
    sI = 1.4
    mT = 0.90
  }
  
  g2 = ggplot(scDat, aes(x = reorder(clone, PL_rank), y = prediction,fill = cellLine)) + 
    scale_fill_manual("Legend", values = dd.col) + 
    geom_violin() + 
    geom_signif(data=scDat,comparisons = subpopComps$clones, map_signif_level = mapSigLevel,step_increase=sI, test = "t.test",margin_top = mT) +    stat_summary(fun = median,geom = "point",size = 2,color = "black") +
    ylab(axesTitleList[param]) + xlab(NULL) + 
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 14),axis.text.y = element_text(size = 12), legend.position = "none", axis.title = element_text(size = 18))
 
  if(param=="K"){
    g3 = g2 + coord_trans( y = "log10");
  }
  
  # produce parameter over biomarker plot
  fitSummaryTable$biomarkerActivity = as.data.frame(grpstats(as.matrix(uList$pathways[inputList[[param]],]), uList$clMembership, "median"))[rownames(fitSummaryTable),]
  fitSummaryTable$parameterValue = fitSummaryTable[,param]
  fitSummaryTable$mad = sapply(fitSummaryTable$cellLine, function(x) mad(scDat[scDat$cellLine==x, "indepVar1"]))
  gg1 = ggplot(fitSummaryTable, aes(x=parameterValue, y=biomarkerActivity, group = cellLine, color = cellLine)) + 
    scale_color_manual("Legend", values = dd.col) + 
    geom_line() + 
    geom_point(size=9, shape=c(16,16,10,16,16,10,10,16)) + 
    geom_errorbar(aes(ymin=biomarkerActivity-mad, ymax=biomarkerActivity+mad)) +
    xlab(axesTitleList[param]) + ylab(inputList[[param]]) + theme_minimal() + theme(axis.title = element_text(size = 18), axis.text = element_text(size = 14), legend.position = "none")
  if(param=="K"){
    gg2 = gg1 + scale_x_log10()
  }
  
  # gather pairs of clones with significant difference
  compsList = getLargestSignificantPair(scDat, clonesToCellLinesMap,cellLinePairs,maxPval=0.1,maxOnly = F)
  pairs[[param]]=compsList$subpopComps
  colnames(pairs[[param]])=paste0(param,"_",colnames(pairs[[param]]))
}

print(gg1)
print(gg2)
print(g2)
print(g3)

## check out subpop growth params
cloneCounts = plyr::count(unlist(sapply(subpopGrowthParams, function(x) rownames(x), simplify = F)))
clones= cloneCounts$x
subpopGrowthParams = base::cbind(subpopGrowthParams$K[clones,,drop=F], subpopGrowthParams$r[clones,,drop=F])
colnames(subpopGrowthParams) = c("K", "r")
subpopGrowthParams$CL = clonesToCellLinesMap[rownames(subpopGrowthParams), "cellLine"]
print(subpopGrowthParams)
```

## Generate Figure 1B
Print the top KEGG pathway biomarker for r/K params across cell lines identified in previous section.
```{r}
## Validation set: fewer pathways but tested for all cell lines:
sT=sapply(teList, function(x) cbind(rownames(x),x), simplify = F)
pTable=do.call(cbind,sT)
## Indices for top K pathways
kI=order(pTable$K.p.value)[1:5];
kI=teList$K[pTable$`K.rownames(x)`[kI],]
## Indices for top r pathways
rI=order(pTable$r.p.value)[1:5]; 
rI=teList$r[pTable$`r.rownames(x)`[rI],]
## p-val adjust
rK_Validation=rbind(rI,kI)
rK_Validation$correctedPval=p.adjust(rK_Validation$p.value,method = "fdr")
print(rK_Validation[,c(1,3:4)])
```

## Generate Figure 1E
The required input to this section ('cloneMembershipBarplot') is gathered in the same section that generates Figures 1C-D, and is hard-coded to display cells in G0/G1 phase which passed QC. This can be changed to display other cell cycle states or all cells.
```{r}
## produce stacked bar plots for clonal membership
## number of clones may change depending on which cell cycle state you are looking at 
new.dd.col=c("#FFF5EB","#F16913", "#FEE6CE", "#FDD0A2", "#FDAE6B" ,"#FD8D3C", "#D94801", "#A63603","#4292C6","#2171B5","#08519C", "#08306B", "#FB6A4A", "#EF3B2C", "#A50F15","#CB181D", "#67000D", "#D7301F", "#B30000","#FFF5F0", "#FEE0D2","#FCBBA1", "#FC9272","#C7E9C0", "#A1D99B","#41AB5D", "#238B45", "#006D2C", "#00441B", "#FFFFCC", "#FED976","#FFEDA0","#F768A1","#FCC5C0", "#FA9FB5","#DEEBF7","#F7FBFF", "#C6DBEF" ,"#9ECAE1") 
names(new.dd.col) = cloneMembershipBarplot$clone
ggplot(data=cloneMembershipBarplot, aes(x=" ", y=ratio, group=cellLine, color=cellLine, fill=clone)) +
  scale_fill_manual("Legend", values = new.dd.col) +
  scale_color_manual("Legend", values = dd.col[unique(cloneMembershipBarplot$cellLine)]) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) + 
  facet_grid(.~ cellLine) + theme_void() + theme(legend.position="none")
```

## Generate Figure 2A
The required input to this section ('pairs') is gathered in the same section that generates Figures 1C-D. It contains the pairs of clones within a cell line that contain significantly different r/K parameters (P<0.1). 
```{r}
## Find clone pairs with potential r/K trade-offs:
rK_pairs=cbind(pairs$r,pairs$K[rownames(pairs$r),])
rK_pairs=rK_pairs[!is.na(rK_pairs$r_src) & !is.na(rK_pairs$K_src),]
estimate=as.numeric(rbind(rK_pairs$r_estimate, rK_pairs$K_estimate))
pval=as.numeric(rbind(rK_pairs$r_pval, rK_pairs$K_pval))
## Plot them
col=sort(rep(dd.col[rK_pairs$r_src],2), decreasing = TRUE)
ymax=max(estimate)*1.2
barplot(estimate, col=col, names=rep(c("r","K"),nrow(rK_pairs)), ylab = "clone 1 / clone 2", ylim=c(0,ymax))
points(c(-1,100),c(1,1),type = "l")
leg=paste0(rK_pairs$r_src,": ", sapply(rK_pairs$r_clones, paste, collapse=","))
legend("topright", leg, fill=unique(col),title = "clone pairs",inset = c(0.05,0.1))
text(seq(0.75,length(pval)*1.15,1.15),rep(ymax*0.96,4),round(pval,4))
```